---
title: "Project 3"
author: "Nash Lawrence & Kevin Wlosinski"
date: "10/17/2022"
output: pdf_document
---


```{r setup, include=FALSE}
library(ResourceSelection)
library(ggmosaic)
library(ggplot2)
library(caret)
library(gains)
library(rpart)
library(rpart.plot)
library(pROC)
library(tidyverse)
library(formatR)
library(MASS)
library(boot)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE)
options(tibble.width = Inf) # displays all columns.
options(tibble.print_max = Inf) # to show all the rows.
```


### Read the data set


```{r}
idTheft <- read_csv("C:/Users/stewi/BUS 458H/StolenRecords.csv")
head(idTheft)
```


### Split the data set into a training and a validation data set

```{r}
# First change the response variable to a nominal.
idTheft$IDTheftVictim <- as.factor(idTheft$IDTheftVictim)
set.seed(1)
myIndex <- createDataPartition(idTheft$IDTheftVictim, p=0.7, list=FALSE)
trainSet <- idTheft[myIndex,]
validationSet <- idTheft[-myIndex,]
```


### Build a couple of mosaic plots for illustration

```{r}
# Effect of using strong passwords

z <- prop.table(table(trainSet$UseStrongPass, trainSet$IDTheftVictim), margin=1) * 100
    rownames( z ) <- c("Not a Strong Passwd","Strong Passwd")
    colnames( z) <- c( "Not ID Theft","ID Theft")
    print( z)
ggplot(data = trainSet) +
  geom_mosaic(aes(x = product(UseStrongPass), fill=IDTheftVictim)) +
  theme_mosaic()

# Effect of direct deposit
z <- prop.table(table(trainSet$DirectDeposit, trainSet$IDTheftVictim), margin=1) * 100
    rownames( z ) <- c("No Direct Deposit","Direct Deposit")
    colnames( z) <- c( "Not ID Theft","ID Theft")
    print( z)
ggplot(data = trainSet) +
  geom_mosaic(aes(x = product(DirectDeposit), fill=IDTheftVictim)) +
  theme_mosaic()

# Effect of ATM card
z <- prop.table(table(trainSet$ATMCard, trainSet$IDTheftVictim), margin=1) * 100
    rownames( z ) <- c("No ATM Card","ATM Card")
    colnames( z) <- c( "Not ID Theft","ID Theft")
    print( z)
ggplot(data = trainSet) +
  geom_mosaic(aes(x = product(ATMCard), fill=IDTheftVictim)) +
  theme_mosaic()

# Effect of Debit Card
z <- prop.table(table(trainSet$DebitCard, trainSet$IDTheftVictim), margin=1) * 100
    rownames( z ) <- c("No Debit Card","Debit Card")
    colnames( z) <- c( "Not ID Theft","ID Theft")
    print( z)
ggplot(data = trainSet) +
  geom_mosaic(aes(x = product(DebitCard), fill=IDTheftVictim)) +
  theme_mosaic()

# Effect of preauthorized debts
z <- prop.table(table(trainSet$PreauthorizedDebts, trainSet$IDTheftVictim), margin=1) * 100
    rownames( z ) <- c("No Preauthorized debts","Preauthorized debts")
    colnames( z) <- c( "Not ID Theft","ID Theft")
    print( z)
ggplot(data = trainSet) +
  geom_mosaic(aes(x = product(PreauthorizedDebts), fill=IDTheftVictim)) +
  theme_mosaic()

# Effect of automated phone systems
z <- prop.table(table(trainSet$AutomatedPhoneSystem, trainSet$IDTheftVictim), margin=1) * 100
    rownames( z ) <- c("No Automated Phone System","Automated Phone System")
    colnames( z) <- c( "Not ID Theft","ID Theft")
    print( z)
ggplot(data = trainSet) +
  geom_mosaic(aes(x = product(AutomatedPhoneSystem), fill=IDTheftVictim)) +
  theme_mosaic()

# Effect of Computer Banking
z <- prop.table(table(trainSet$ComputerBanking, trainSet$IDTheftVictim), margin=1) * 100
    rownames( z ) <- c("No Computer Banking","Computer Banking")
    colnames( z) <- c( "Not ID Theft","ID Theft")
    print( z)
ggplot(data = trainSet) +
  geom_mosaic(aes(x = product(ComputerBanking), fill=IDTheftVictim)) +
  theme_mosaic()

# Effect of in-person Banking
z <- prop.table(table(trainSet$InPersonBanking, trainSet$IDTheftVictim), margin=1) * 100
    rownames( z ) <- c("No in-person Banking","In-Person Banking")
    colnames( z) <- c( "Not ID Theft","ID Theft")
    print( z)
ggplot(data = trainSet) +
  geom_mosaic(aes(x = product(InPersonBanking), fill=IDTheftVictim)) +
  theme_mosaic()

# Effect of by Mail Banking
z <- prop.table(table(trainSet$ByMailBanking, trainSet$IDTheftVictim), margin=1) * 100
    rownames( z ) <- c("No by-mail Banking","By-mail Banking")
    colnames( z) <- c( "Not ID Theft","ID Theft")
    print( z)
ggplot(data = trainSet) +
  geom_mosaic(aes(x = product(ByMailBanking), fill=IDTheftVictim)) +
  theme_mosaic()

# Effect of Phone Banking
z <- prop.table(table(trainSet$PhoneBanking, trainSet$IDTheftVictim), margin=1) * 100
    rownames( z ) <- c("No Phone Banking","Phone Banking")
    colnames( z) <- c( "Not ID Theft","ID Theft")
    print( z)
ggplot(data = trainSet) +
  geom_mosaic(aes(x = product(PhoneBanking), fill=IDTheftVictim)) +
  theme_mosaic()

# Effect of Changing Passwords Regularly
z <- prop.table(table(trainSet$ChgPassReg, trainSet$IDTheftVictim), margin=1) * 100
    rownames( z ) <- c("No Freq Pass Change","Frequent Password Change")
    colnames( z) <- c( "Not ID Theft","ID Theft")
    print( z)
ggplot(data = trainSet) +
  geom_mosaic(aes(x = product(ChgPassReg), fill=IDTheftVictim)) +
  theme_mosaic()

# Effect of Monitoring Bank Statements Regularly
z <- prop.table(table(trainSet$MonitorBankStmt, trainSet$IDTheftVictim), margin=1) * 100
    rownames( z ) <- c("No Freq Monitor Bank","Freq Monitors Bank")
    colnames( z) <- c( "Not ID Theft","ID Theft")
    print( z)
ggplot(data = trainSet) +
  geom_mosaic(aes(x = product(MonitorBankStmt), fill=IDTheftVictim)) +
  theme_mosaic()

# Effect of Monitoring Credit Card Statements
z <- prop.table(table(trainSet$MonitorCCStmt, trainSet$IDTheftVictim), margin=1) * 100
    rownames( z ) <- c("No Freq Monitor CC","Freq Monitors CC")
    colnames( z) <- c( "Not ID Theft","ID Theft")
    print( z)
ggplot(data = trainSet) +
  geom_mosaic(aes(x = product(MonitorCCStmt), fill=IDTheftVictim)) +
  theme_mosaic()

# Effect of Monitoring Credit Card Reports Regularly
z <- prop.table(table(trainSet$MonitorCredreport, trainSet$IDTheftVictim), margin=1) * 100
    rownames( z ) <- c("No Freq Monitor CR","Freq Monitors CR")
    colnames( z) <- c( "Not ID Theft","ID Theft")
    print( z)
ggplot(data = trainSet) +
  geom_mosaic(aes(x = product(MonitorCredreport), fill=IDTheftVictim)) +
  theme_mosaic()
```

### Interpretation of the mosaic plots
1. Customers who do not use strong passwords are more likely to be victims of id theft than customers who use strong passwords.
2. Customers who use direct deposit are more likely to be victims of id theft than customers who do not use direct deposit.
3. Customers who use an ATM card are more likely to be victims of id theft than customers who do not use an ATM card.
4. Customers who use a debit card are more likely to be victims of id theft than customers who do not use a debit card.
5. Customers who use pre-authorized debts are more likely to be victims of id theft than customers who do not use pre-authorized debts.
6. Customers who use the automated phone system are slightly more likely to be victims of id theft than customers who do not use the automated phone system.
7. Customers who use computer banking are more likely to be victims of id theft than customers who do not use computer banking.
8. Customers who do not use in-person banking are more likely to be victims of id theft than customers who do use in-person banking.
9. Customers who do not use by-mail banking are more likely to be victims of id theft than customers who do use by-mail banking.
10. Customers who do not use phone banking are more likely to be victims of id theft than customers who do use phone banking.
11. Customers who do not change their password regularly are more likely to be victims of id theft than customers who do change their password regularly.
12. Customers who do not monitor their bank statement are more likely to be victims of id theft than customers who do monitor their bank statement.
13. Customers who do not monitor their credit card statement are more likely to be victims of id theft than customers who do monitor their credit card statement.
14. Customers who do not monitor their credit report are more likely to be victims of id theft than customers who do monitor their credit report.



### Build the logistic regression model.

```{r}
LR_Model<-glm(IDTheftVictim ~ DirectDeposit + ATMCard	+ DebitCard + PreauthorizedDebts + AutomatedPhoneSystem + ComputerBanking + InPersonBanking + ByMailBanking	+ PhoneBanking + ChgPassReg + UseStrongPass + MonitorBankStmt	+ MonitorCCStmt + MonitorCredreport, data=trainSet, family=binomial)
summary(LR_Model)
exp(coef(LR_Model))
hl <- hoslem.test(trainSet$IDTheftVictim, fitted(LR_Model), g=10)
hl
```
### Interpretation of the odds ratios
1. Holding all other variables constant, customers who use strong passwords are 67% less likely to be victims of id theft than customers who do not use strong passwords.
2. Holding all other variables constant, customers who use direct deposit are 73% more likely to be victims of id theft than customers who do not use direct deposit.
3. Holding all other variables constant, customers who use an ATM card are 114% more likely to be victims of id theft than customers who do not use an ATM card.
4. Holding all other variables constant, customers who use a debit card are 35% more likely to be victims of id theft than customers who do not use a debit card.
5. Holding all other variables constant, customers who use preauthorized debts are 187% more likely to be victims of id theft than customers who do not use preauthorized debts.
6. Holding all other variables constant, customers who use the automated phone system are 5% more likely to be victims of id theft than customers who do not use the automated phone system.
7. Holding all other variables constant, customers who use computer banking are 330% more likely to be victims of id theft than customers who do not use computer banking. 
8. Holding all other variables constant, customers who use in-person banking are 40% less likely to be victims of id theft than customers who do not use in-person banking.
9. Holding all other variables constant, customers who use by-mail banking are 18% less likely to be victims of id theft than customers who do not use by-mail banking.
10. Holding all other variables constant, customers who use phone banking are 25% less likely to be victims of id thefft than customers who do not use phone banking.
11. Holding all other variables constant, customers who change their password regularly are 70% less likely to be victims of id theft than customers who do not change their password regularly.
12. Holding all other variables constant, customers who monitor their bank statement are 29% less likely to be victims of id theft than customers who do not monitor their bank statement.
13. Holding all other variables constant, customers who monitor their credit card statement are 46% less likely to be victims of id theft than customers who do not monitor their credit card statement.
14. Holding all other variables constant, customers who monitor their credit report are 54% less likely to be victims of id theft than customers who do not monitor their credit report.


### Predict the validation data set

```{r}
predicted_prob <- predict(LR_Model, validationSet, type="response")

# Let's get the confusion matrix for the validation data set
validationSet<- validationSet %>% mutate(predicted_class = ifelse(predicted_prob > 0.5, 1, 0))
table(validationSet$predicted_class, validationSet$IDTheftVictim)

# Let's get the confusion matrix for the training data set
predicted_prob1 <- predict(LR_Model, trainSet, type="response")
trainSet <- trainSet %>% mutate(predicted_class = ifelse(predicted_prob1 > 0.5, 1, 0))
table(trainSet$predicted_class, trainSet$IDTheftVictim)
# Note: Rows are the predicted classes while the columns are the observed classes
```

### Confusion Matrix Interpretation
Sensitivity = 18322 / (39497 + 18322) = .3169 = 32% of individuals who are victims of id theft are also predicted to be victims of id theft
Specificity = 231863 / (231863 + 39497) = .8544 = 85% of individuals who are not victims of id theft are also predicted to not be victims of id theft
Misclassification Rate = (10318 + 39497) / (300000) = .16605 = 17% of individuals are incorrectly predicted to be victims or non-victims, when they truly belong to the opposite category.

The model is good at predicting at non-victims.


### Create the gains table

```{r}
validationSet$IDTheftVictim<-as.numeric(as.character(validationSet$IDTheftVictim))
gains_table<-gains(validationSet$IDTheftVictim, predicted_prob)
gains_table
```

### Interpretation of Lift at 10%
The top 10%, based on predicted probability, are 3.28 times more likely to be identified as victims of id theft than a randomly selected individual.


### Create the plots

```{r}
plot(c(0,gains_table$cume.pct.of.total*sum(validationSet$IDTheftVictim))~
       c(0,gains_table$cume.obs), xlab="# of cases",
     ylab="Cumulative",main="Cumulative Lift Chart",type="l")
lines(c(0,sum(validationSet$IDTheftVictim))~c(0,dim(validationSet)[1]),col="red",lty=2)
barplot(gains_table$mean.resp/mean(validationSet$IDTheftVictim),
        names.arg=gains_table$depth, xlab="Percentile", ylab="Lift",
        ylim=c(0,3), main="Decile-Wise Lift Chart")
roc_object<-roc(validationSet$IDTheftVictim,predicted_prob)
plot.roc(roc_object)
auc(roc_object)
```

### Interpretation of C-statistic
By using this model, we are 82% accurate in discriminating between victims and non-victims of id theft.


## Stepwise Selection - Start with full model

```{r}
step.model <- LR_Model %>% stepAIC(trace = FALSE)
summary(step.model)
```

## Backward Selection - Start with full model

```{r}
back.model <- LR_Model %>% stepAIC(direction = "backward", trace = FALSE)
summary(back.model)
```

## Forward Selection - Software does its job. It starts with the intercept only model and then builds the final model.

```{r}
forward.model <- LR_Model %>% stepAIC(direction = "forward", trace = FALSE)
summary(back.model)
```

## K-Fold Logistic Regression Model

```{r}
fitcontrol1 <- trainControl(method="cv", number = 5, savePredictions = T)
mod_fitcv <- train(IDTheftVictim~DirectDeposit+ATMCard	+DebitCard+PreauthorizedDebts+AutomatedPhoneSystem	+ComputerBanking+InPersonBanking+ByMailBanking	+PhoneBanking+ChgPassReg+UseStrongPass+MonitorBankStmt	+MonitorCCStmt+MonitorCredreport
, data=idTheft, method = "glm", family="binomial", trControl = fitcontrol1)
summary(mod_fitcv)
```

## Another K-Fold Code

```{r}
credit_glm1<- glm(IDTheftVictim~DirectDeposit+ATMCard	+DebitCard+PreauthorizedDebts+AutomatedPhoneSystem	+ComputerBanking+InPersonBanking+ByMailBanking	+PhoneBanking+ChgPassReg+UseStrongPass+MonitorBankStmt	+MonitorCCStmt+MonitorCredreport, family=binomial, data=idTheft);  
cv_result  <- cv.glm(data=idTheft, glmfit=credit_glm1, K=10) 
cv_result$delta[1]
```



