---
title: "Final Project"
author: "Nash Lawrence"
date: "4/20/22"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Install and load the packages needed for this assignment

```{r echo = FALSE}
# list of packages require for this assignment
packages = c("tidyverse", "ggplot2", "readxl")

## now load or install and load
package.check <- lapply(
  packages,
  FUN = function(x) {
    if (!require(x, character.only=TRUE)) {
      install.packages(x, dependencies=TRUE)
      library(x, character.only=TRUE)
    }
  }
)
```
## 1.) Give description of your dataset including a description of the variables

My dataset is a CSV file I made back in March to analyze all 68 teams in the NCAA Men's Basketball Tournament

1. team_id: each teams unique numeric code numbered 1-68
2. team_name: the name of each school, some abbreviated (Ex-"UNC" for North Carolina University)
3. team_conference: the abbreviation of the conference the school is in (Ex- "ACC" or "Pac 12")
4. p6_conference: a boolean to see if the school is in one of the main 6 conferences(ACC/Big East/Big Ten/Big 12/Pac 12/SEC). Y= Yes, N= No
5. conference_champ: a boolean to see if the school won their conference tournament championship. Y=Yes, N=No
6. tourney_seed: the number seed the school is in their respective region. Integer 1-16, 1=Best rank, 16=Worst rank
7. tourney_region: the regional bracket each school is in(West/South/East/Midwest)
8. opponent_name: the name of the school that each team is playing in their first tournament game
9. total_wins: the total amount of wins the school had on the season overall
10. total_loss: the total amount of losses the school had on the season overall
11. conference_wins: the total amount of wins the school had against teams in their conference
12. conference_loss: the total amount of losses the school had against teams in their conference
13. home_wins: the total amount of wins the school had in home games
14. home_loss: the total amount of losses the school had in home games
15. road/n_wins: the total amount of wins the school had in away or neutral site games
16. road/n_loss: the total amount of losses the school had in away or neutral site games
17. SOS: the schools ranking in Strength of Schedule (Rank 1-354)
18. SOR: the schools ranking in the Strength of Record metric which measures how impressive a teams record is based on opponents
19. KP: the schools ranking in KenPom, a system that ranks teams based on a number of metrics and statistics
20. NET: the schools ranking in the NCAA Evaluation Tool, a system that ranks teams based on metrics and statistics
21. off_efficiencyRank: the schools ranking in offensive efficiency, which is points scored per possession
22. def_efficiencyRank: the schools ranking in defensive efficiency, which is points allowed per possession
23. pace_rank: the schools ranking in possessions per game to measure which teams play at a fast/slow pace
24. team_ppg: the schools points per game average
25. fg_percentage: the schools overall shooting percentage
26. three_percentage: the schools overall shooting percentage on three point shots
27. ft_percentage: the schools overall shooting percentage on free throw shots
28. 3PA_perGame: the schools average for three point shots attempted each game
29. assists_perGame: the schools assists per game average
30. rebounds_perGame: the schools rebounds per game average
31. offReb_perGame: the schools offensive rebounds per game average
32. steals_pg: the schools steals per game average
33. blocks_pg: the schools blocks per game average
34. turnovers_pg: the schools turnovers per game average (fewer = better)


## 2.) Load your dataset and save it to a RTibble/Pandas dataframe object and display the first 5 rows

```{r}
url <- "C:/Users/stewi/BUS440/Personal Projects/teamData.csv"
teams <- read_csv(url)
teams %>% head()
```

## 3.) Identify the numerical and categorical columns you will investigate. Write a paragraph describing why you chose them.

The categorical variables I am going to use are: team_name, team_conference, p6_conference, tourney_region.
The numerical variables I am going to use are: SOR, off_efficiencyRank, def_efficiencyRank, team_ppg.

I am using team_name to identify each team when comparing data, I am using team_conference to compare data between different conferences. I am using p6_conference to compare data between teams in major conferences and teams in mid-major/small conferences. I am using tourney_region to compare data between teams in each region since every region has a team seeded 1-16.

I am using SOR because strength of record is a great metric which takes plenty of things into account when ranking how impressive a teams resume is. I am using off_efficiency because it is a great statistic to determine how often a team scores and not just how many points they score total (because some teams have less possessions per game). I am also using def_efficiency because it is a great statistic to determine how often a team allows their opponent to score. I am using team_ppg because it shows which teams score the most points per game and it is a popular statistic.

## 4.) Create some numerical summaries and display the results in a table
***Number of P6 conferences and non-P6 conferences per region***

```{r}
numeric_summary <- table(teams$p6_conference, teams$tourney_region)
numeric_summary
```

***Find the number of teams that each conference has for each seed number***

```{r}
numeric_tbl <- table(teams$team_conference, teams$tourney_seed)
numeric_tbl
```

## 5.) Include measures of center (mean, median) and spread (standard deviation) across a single grouping variable
***Strength of Record (SOR) for each of the Power 6 conferences***

```{r}
SOR_measures <- teams %>%
  filter(p6_conference == "Y") %>%
  group_by(team_conference) %>%
  select(team_conference, SOR) %>%
  summarise("SOR Average" = mean(SOR, na.rm = TRUE), "SOR Median" = median(SOR, na.rm = TRUE),
            "SOR Standard Deviation" = sd(SOR, na.rm = TRUE),) %>%
  rename("Conference" = team_conference)

SOR_measures
```

## 6.) Create a new categorical column based on a numerical column
***25 years straight the national champion has been top 40 in offensive efficiency and top 25 in defensive efficiency***

```{r}
mutate_contender <- teams %>%
  select(team_name) %>%
  mutate(Contender_Frauds = ifelse ((teams$off_efficiencyRank < 41 & teams$def_efficiencyRank < 26),
                                      "Contender", "Frauds")) %>%
  rename("Team" = team_name)

mutate_contender
```

## 7.) Create a bar plot for two categorical variables. For one use the column created in Q6

```{r}
mutated_bar <- ggplot(data = mutate_contender, (aes(x = Contender_Frauds)))
mutated_bar + geom_bar(color = "black", fill = "red", size = 2, bindwidth = 1) + labs(x = "Contenders/Frauds", y = "Count of Teams\n", title = "Number of Teams in each Classification based on Metrics") + theme(plot.title = element_text(hjust = 0.5)) + scale_y_continuous(breaks=c(0,10,20,30,40,50,60))
```

***bar plot of the number of tournament teams in each of the Power 6 conferences***
```{r}
num_teams <- teams %>%
  filter(p6_conference == "Y") %>%
  select(team_name, team_conference) %>%
  rename("Conference" = team_conference, "Team" = team_name)
```

```{r}
p6_conferenceReps <- ggplot(data = num_teams, (aes(x = Conference, fill = Conference)))
p6_conferenceReps + geom_bar(color = "black", size = 1.5, bindwidth = 1.5) + labs(y = "Number of Tournament Teams\n", x = "", title = "Number of Tournament Teams per Major Conference") + theme(plot.title = element_text(hjust = 0.5)) + scale_y_continuous(breaks=c(0,1,2,3,4,5,6,7,8,9))
```

## 8.) Create a histogram for a numeric variable
```{r}
teamPoints_histo <- ggplot(data = teams, (aes(x = team_ppg)))
teamPoints_histo + geom_histogram(color = "black", fill = "dark green", size = 1.2, bindwidth = 1, bins = 5) + labs(x = "Points Per Game", y = "Number of Teams per Interval\n", title = "Team Points Per Game Averages") + theme(plot.title = element_text(hjust = 0.5)) + scale_x_continuous(breaks=c(65,70,75,80,85,90)) + scale_y_continuous(breaks=c(0,5,10,15,20,25,30,35))
```

## 9.) Create an overlayed histogram for a numeric variable across one of the categorical variables

```{r}
ppg <- teams$team_ppg
p6 <- teams$p6_conference

overlay_histo <- ggplot(data = teams, (aes(x = ppg, fill = p6)))
overlay_histo + geom_histogram(position = "identity", alpha = 0.5, color="black", bins = 20) + labs(x = "Points Per Game Averages\n", y = "Count of Teams\n", title = "Team Points Per Game Averages for Power 6/Non-Power Conferences", fill = "Major Conference") + theme(plot.title = element_text(hjust = 0.5)) + scale_x_continuous(breaks=c(65,70,75,80,85,88)) + scale_y_continuous(breaks=c(0,1,2,3,4,5,6,7,8,9,10,11))
```

## 10.) Create a scatter plot relating two numeric variables including the least squares regression line

```{r}
scat_plot <- ggplot(data = teams, aes(x = off_efficiencyRank, y = def_efficiencyRank))
scat_plot + geom_point(aes(color = as.factor(tourney_seed))) + geom_smooth(method = "lm") + labs(x = "Offensive Efficiency Rank\n", y = "Defensive Efficiency Rank\n", title = "Efficiency Ranking Correlations", color = "Seed Number") + theme(plot.title = element_text(hjust = 0.5)) + scale_x_continuous(breaks=c(1,50,100,150,200,250,290)) + scale_y_continuous(breaks=c(1,50,100,150,200,250,275))
```

***Correlation Coefficient***

```{r}
cor_coeff <- cor(teams$off_efficiencyRank, teams$def_efficiencyRank)
cor_coeff
if (cor_coeff > 0) {
  print('The correlation between the offensive and defensive efficiencies of teams is positive')
} else {
  print('The correlation between the offensive and defensive efficiencies of teams is negative')
}
```


## 11.) Use sckit learn to find the slope and y-intercept of the least squares regression line. Plot the line on the scatter plot.

```{r}
efficiency_model <- lm(off_efficiencyRank ~ def_efficiencyRank, data = teams)
efficiency_model
```








