---
title: "Project 2 BUS 458H"
author: "Nash Lawrence & Kevin Wlosinski"
date: "9/18/2022"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)
```


***2.) Read the data and provide Summary Statistics for both variables. Interpretations not needed***

```{r}
url <- "C:/Users/stewi/BUS 458H/Project_2.csv"
projectData <- read_csv(url)
head(projectData)
```

```{r}
summary(projectData$MPG_City)
summary(projectData$Horsepower)
```


***3.) Construct a scatter plot between MPG_City and Horsepower. Ensure that this plot is labeled correctly. Thoroughly interpret the plot. You may complete your write-up under the section headers.***


```{r}
scatPlot <- ggplot(data = projectData, aes(x = MPG_City, y = Horsepower))
scatPlot + labs(x = "Miles per Gallon (City)\n", y = "Horsepower\n", title = "Miles per Gallon vs Horsepower") + theme(plot.title = element_text(hjust = 0.5)) + scale_x_continuous(breaks = c(10,20,30,40,50,60)) + geom_point() + stat_smooth(method = lm, formula = y ~ x + I(x^2))
```

## Interpretations: The higher the Horsepower, the lower the MPG in the City. The graph seems to be exponential based on the pattern of the values on the graph. The two data points around 60 MPG look to be outliers, as all but 3 values are below 40 on the X-Axis. We counted 7 outliers for the MPG_City variable, and Zero for the Horsepower variable.


***Correlation between variables***

```{r}
cor(projectData$MPG_City, projectData$Horsepower)
```

***Number of outliers for the MPG variable***

```{r}
Qone <- quantile(projectData$MPG_City, 0.25)
Qthree <- quantile(projectData$MPG_City, 0.75)
IQR <- Qthree - Qone
LOF <- Qone - 3.0 * IQR
UOF <- Qthree + 3.0 * IQR
numOutliesX <- projectData %>%
  filter(MPG_City < LOF | MPG_City > UOF)

count(numOutliesX)
```

***Number of outliers for Horsepower variable***

```{r}
QoneY <- quantile(projectData$Horsepower, 0.25)
QthreeY <- quantile(projectData$Horsepower, 0.75)
IQRy <- QthreeY - QoneY
LOFy <- QoneY - 3.0 * IQRy
UOFy <- QthreeY + 3.0 * IQRy
numOutliesY <- projectData %>%
  filter(Horsepower < LOFy | Horsepower > UOFy)

count(numOutliesY)
```





***4.) Conduct a simple linear regression between MPG_City and HP. Interpret the coefficient associated with Horsepower. Also interpret the R-Square value.***

```{r}
linReg <- lm(MPG_City ~ Horsepower, data = projectData)
linReg
```

## The coefficients suggest that Miles Per Gallon in the City is predicted by an intercpet of 30.71347, which decreases by an average of 0.04934 per Horsepower. The small standard error value suggests this is statistically significant.

```{r}
summary(linReg)
```
## R-Squared value is 0.4579, meaning 45.79% of the variance in Horsepower is explained by Miles Per Gallon in the City. This is a decent amount of variance explained, but not a great amount. The Adjusted R-Squared value is basically the same, meaning the model is not overfit.

***Confidence intervals for the mean response***

```{r}
predict(linReg, data.frame(Horsepower=c(10,20,30,40,50,60)), interval = "confidence")
```

***Prediction intervals***

```{r}
predict(linReg, data.frame(Horsepower = c(10,20,30,40,50,60)), interval = "prediction")
```



***5.) Conduct a residual analysis for the model built in Step 4. Check with your instructor on the R code needed to accomplish this. Is the model you built in Step 4 appropriate? Why or Why not?***

## The model built in Q4 was appropriate because I was able to conduct linear regression with the two variables, get all of the coefficients and summary statistics, and then get the confidence intervals and prediction intervals

```{r}
resAnalysis <- residuals(linReg)
head(resAnalysis)
```

***Residual Analysis of Original Linear Model***

```{r}
plot(linReg, which = 1)
```

```{r}
plot(linReg, which = 2)
```

```{r}
plot(linReg, which = 3)
```


***Graph of the residual values and their distance from the least squares line***

```{r}
projectData$predicted <- predict(linReg)
ggplot(data = projectData, aes(x = MPG_City, y = Horsepower)) + geom_smooth(method = "lm", se = FALSE, color = "lightgrey") + geom_segment(aes(xend = MPG_City, yend = predicted), alpha = .2) + geom_point(aes(color = abs(resAnalysis), size = abs(resAnalysis))) + scale_color_continuous(low = "green", high = "red") + guides(color = FALSE, size = FALSE) + geom_point(aes(y = predicted), shape = 1) + labs(x = "Miles per Gallon (City)\n", y = "Horsepower\n", title = "Miles per Gallon vs Horsepower: Residual Analysis") + theme(plot.title = element_text(hjust = 0.5)) + scale_x_continuous(breaks = c(10,20,30,40,50,60)) + scale_y_continuous(breaks = c(100,200,300,400,500))
```

***Statistical summary of the residual values***

```{r}
summary(resAnalysis)
```

***Find the Residual Sums of Squares (RSS)***

```{r}
squareResids <- round(resAnalysis ** 2, digits = 3)
RSS <- sum(squareResids)
RSS
```



***6.) Repeat Step 4 by log-transforming MPG_City. Interpret the coefficient associated with Horsepower.***

```{r}
projectData$mpgLog <- log(projectData$MPG_City)
head(projectData$mpgLog)
```

```{r}
simpReg_logX <- lm(mpgLog ~ Horsepower, data = projectData)
simpReg_logX
```

```{r}
summary(simpReg_logX)
```

## Interpretations: After you log transform the MPG_City variable, the predicted value of MPG_City is calculated by an intercept of 3.4731752 which decreases by 0.0023237 per Horsepower



***7.) Repeat Step 4 by log-transforming MPG_City as well as Horsepower. Interpret the coefficient associated with Horsepower.***

```{r}
projectData$HorseLog <- log(projectData$Horsepower)
head(projectData$HorseLog)
```

```{r}
bothLog_linReg <- lm(mpgLog ~ HorseLog, data = projectData)
bothLog_linReg
```

```{r}
summary(bothLog_linReg)
```

## Interpretations: After you log transform both the MPG_City and Horsepower variables, the predicted value of MPG_City is calculated by an intercept of 5.9097 which decreases by 0.5522 per Horsepower



***8.) Conduct a thorough residual analysis for the model you built in Step 7. Is this model appropriate? Why or why not?***

## The model built in step 7 is appropriate because I log transformed both variables and conducted a linear regression model with the log transformed variables to interpret the coefficients

```{r}
analysis_residLog <- residuals(bothLog_linReg)
head(analysis_residLog)
```

```{r}
summary(analysis_residLog)
```

```{r}
plot(bothLog_linReg, which = 1)
```

```{r}
plot(bothLog_linReg, which = 2)
```

```{r}
plot(bothLog_linReg, which = 3)
```


***Get the Residual Sums of Squares (RSS)***

```{r}
squareLog_resids <- round(analysis_residLog ** 2, digits = 3)
RSS_log <- sum(squareLog_resids)
RSS_log
```





